---
title: "NSBB552 Final Project (2021) - Phoebe Chum"
output:
  pdf_document:
    latex_engine: xelatex
  html_notebook: default
  word_document: default
---

# **"Cerebrovascular miRNA Profile in Alzheimer's Disease"**

## Approach

### **Mice Model**

Triple transgenic Alzheimer's disease "3xTg-AD" mouse model with three human genes Amyloid-beta precursor protein (APP), Presenilin 1 (PSEN1), and Microtubule-associated protein tau (MAPT).

### Tissue Type

Whole brain cerebral vessels.

### Study Design

-   Sample groups:

    -   Young control (YC; 1-2 mo)

    -   Cognitive impairment (CI; 4-5 mo)

    -   Amyloid-beta (AB; 6-8 mo)

    -   AB+Tau (ABT; greater than 12 mo).

-   n = 3 males and 3 females for each of the four groups.

-   Total RNA was extracted from the vessels and sent to NanoString for miRNA expression panel.

-   miRNA expression was then analyzed with DESeq2.

### **Hypothesis**

*Cerebrovascular microRNA (miRNA) expression profiles corresponding to post-transcriptional regulation can provide a diagnostic map of the early development of AD pathology.*\
\

## Packages

```{r}
library(DESeq2)
library(pheatmap)
library(gplots)
library(ggplot2)
```

### Resources

Tutorial:

1.  <https://www.youtube.com/watch?v=wPzeea1Do18>

Documentation:

1.  <http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#countmat>

2.  <https://leaherb.com/add-gradient-colors-to-bar-chart-with-ggplot/>

3.  <https://stackoverflow.com/questions/48633302/reordering-bars-in-a-group-with-ggplot-and-adjusting-groups-by-width>

## Input Data

Load count matrix and coldata

```{r}
cts <-read.csv("Raw Data_Formatted.csv", header = T, row.names = 1)

coldata <-read.csv("MetaData.csv", header = T, row.names = 1)
```

```{r}
head(cts,10)
```

```{r}
coldata
```

Build an DESeqDataSet from a count matrix and a table of sample information

```{r}
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~Age + Sex)

```

Pre-filtering the dataset/remove rows with zero count

```{r}
dds <- dds[ rowSums(counts(dds)) > 12, ]
```

Differential expression analysis with

```{r}
ddsDE <-DESeq(dds)
```

Export normalized read counts

```{r}
normCounts <-counts(ddsDE, normalized = T)
write.csv(normCounts, "normal.all4.csv")
```

### Extract result

specify the two groups for comparison

#### MCI vs YC

```{r}
res1 <- results(ddsDE, contrast=c("Age", "MCI", "YC"))
res1
summary(res1)
resultsNames(ddsDE)
```

#### AB vs YC

```{r}
res2 <- results(ddsDE, contrast=c("Age", "AB", "YC"))
res2
summary(res2)
```

#### ABT vs YC

```{r}
res3 <- results(ddsDE, contrast=c("Age", "ABT", "YC"))
res3
summary(res3)
```

## Exploratory Data Analysis (EDA)

```{r}
plotMA(ddsDE)
```

### Data Transformation

```{r}
dds <- estimateSizeFactors(dds)
dds <- estimateDispersions(dds)
vsd <- varianceStabilizingTransformation(dds)
```

### Heatmap

```{r}
library(pheatmap)
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)
df <- as.data.frame(colData(dds)[,c("Age","Sex")])
pheatmap(assay(vsd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=TRUE, annotation_col=df)
```

### PCA

```{r}
library(ggplot2)
pcaData <- plotPCA(vsd, intgroup=c("Age", "Sex"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=Age, shape=Sex)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```

## Log2 Fold Change Comparison

### MCI vs YC

Convert result into data frame, the filter

```{r}
res1 <- as.data.frame(res1)
res1p <- res1[res1$pvalue<0.05,]
res1padj <- res1[res1$padj<0.05,]
res1logUp <- res1[res1$log2FoldChange>=1,]
res1logDown <- res1[res1$log2FoldChange<=-1,]

res1logUp
res1logDown
```

**Comments:**

+-----------------------------+-------------------+--------------+----------------+
| log2 Fold Change Comparison | Criteria          | Up-regulated | Down-regulated |
+=============================+===================+==============+================+
| MCI vs YC                   | Up-regulated:     | 7 miRNAs     | 3 miRNAs       |
|                             |                   |              |                |
| (n=6 v 6)                   | Fold-change \>=1  |              |                |
|                             |                   |              |                |
|                             | Down-regulated:   |              |                |
|                             |                   |              |                |
|                             | Fold-change \<=-1 |              |                |
+-----------------------------+-------------------+--------------+----------------+

```{r}
MCI_log2 <-read.csv("MCI vs YC_log2.csv", header = T, row.names = 1)

ggplot(data = MCI_log2, 
       aes(x = miRNA, y = Log2FoldChange,
           )) + 
  geom_bar(stat = "identity", aes(fill = Log2FoldChange), position=position_dodge()) + 
  ylab("Log2 Fold Change") + 
  coord_flip() + scale_fill_gradient2(low='orange', mid='snow', high='blue')
```

### AB vs YC

Convert result into data frame, the filter

```{r}
res2 <- as.data.frame(res2)
res2p <- res2[res2$pvalue<0.05,]
res2padj <- res2[res2$padj<0.05,]
res2logUp <- res2[res2$log2FoldChange>=1,]
res2logDown <- res2[res2$log2FoldChange<=-1,]

res2logUp
res2logDown
```

**Comments:**

+-----------------------------+-------------------+--------------+----------------+
| log2 Fold Change Comparison | Criteria          | Up-regulated | Down-regulated |
+=============================+===================+==============+================+
| AB vs YC                    | Up-regulated:     | 3 miRNAs     | 1 miRNAs       |
|                             |                   |              |                |
| (n=6 v 6)                   | Fold-change \>=1  |              |                |
|                             |                   |              |                |
|                             | Down-regulated:   |              |                |
|                             |                   |              |                |
|                             | Fold-change \<=-1 |              |                |
+-----------------------------+-------------------+--------------+----------------+

```{r}
AB_log2 <-read.csv("AB vs YC_log2.csv", header = T)

ggplot(data = AB_log2, 
       aes(x = miRNA, y = Log2FoldChange,
           )) + 
  geom_bar(stat = "identity", aes(fill = Log2FoldChange), position=position_dodge()) + 
  ylab("Log2 Fold Change") + 
  coord_flip() + scale_fill_gradient2(low='orange', mid='snow', high='blue')
```

### ABT vs YC

Convert result into data frame, the filter

```{r}
res3 <- as.data.frame(res3)
res3p <- res3[res3$pvalue<0.05,]
res3padj <- res3[res3$padj<0.05,]
res3logUp <- res3[res3$log2FoldChange>=1,]
res3logDown <- res3[res3$log2FoldChange<=-1,]

res3logUp
res3logDown
```

**Comments:**

+-----------------------------+-------------------+--------------+----------------+
| log2 Fold Change Comparison | Criteria          | Up-regulated | Down-regulated |
+=============================+===================+==============+================+
| ABT vs YC                   | Up-regulated:     | 5 miRNAs     | 2 miRNAs       |
|                             |                   |              |                |
| (n=6 v 6)                   | Fold-change \>=1  |              |                |
|                             |                   |              |                |
|                             | Down-regulated:   |              |                |
|                             |                   |              |                |
|                             | Fold-change \<=-1 |              |                |
+-----------------------------+-------------------+--------------+----------------+

```{r}
ABT_log2 <-read.csv("ABT vs YC_log2.csv", header = T)

ggplot(data = ABT_log2, 
       aes(x = miRNA, y = Log2FoldChange,
           )) + 
  geom_bar(stat = "identity", aes(fill = Log2FoldChange), position=position_dodge()) + 
  ylab("Log2 Fold Change") + 
  coord_flip() + scale_fill_gradient2(low='orange', mid='snow', high='blue')
```

### Summary

#### Log2 Fold Change: Up-regulated

| MCI vs YC            | AB vs YC            | ABT vs YC           |
|----------------------|---------------------|---------------------|
| mmu-let-7a           |                     | mmu-miR-183         |
| **mmu-miR-144**      | **mmu-miR-144**     | mmu-miR-1932        |
| mmu-miR-155          | **mmu-miR-291a-5p** | **mmu-miR-291a-5p** |
| mmu-miR-1947         | mmu-miR-690         | mmu-miR-295         |
| mmu-miR-2134         |                     |                     |
| **mmu-miR-429**      |                     | **mmu-miR-429**     |
| [*mmu-miR-543*]{.ul} |                     |                     |

##### Literature Search

-   Elevated Levels of miR-144-3p Induce Cholinergic Degeneration by Impairing the Maturation of NGF in Alzheimer's Disease (doi: 10.3389/fcell.2021.667412)

-   Knockdown of miR-429 Attenuates Aβ-Induced Neuronal Damage by Targeting SOX2 and BCL2 in Mouse Cortical Neurons (doi: 10.1007/s11064-018-2643-3)

#### Log2 Fold Change: Down-regulated

| MCI vs YC      | AB vs YC                 | ABT vs YC                |
|----------------|--------------------------|--------------------------|
| mmu-miR-1939   |                          |                          |
| mmu-miR-380-3p | [***mmu-miR-543***]{.ul} | [***mmu-miR-543***]{.ul} |
| mmu-miR-744    |                          | mmu-miR-720              |

##### Literature Search

-   miR-543: PubMed search - 211 results

## Conclusion

### Observation

For this project, I used DESeq2 to analyze NanoString miRNA count data. The data did not show clear clustering with heatmap and PCA plot. PC1 shows 32% variance and PC2 shows 10% variance. Neither of the principle component suggests a strong distinction to separate the sample groups. Log2 fold change suggests some upregulated and downregulated miRNAs. In particular, I was interested in any miRNAs expressions with at least 2-fold differences. According to the summary table shown above, mmu-miR-144 marks the transition between mild cognitive impairment (MCI) and Amyloid-beta (Aβ) stage whereas mmu-miR-429 marks the transition between Aβ and Amyloid-beta + Tau (AβT)stage. Interenstingly, mmu-miR-543 is upregulated in MCI, but is downregulated in Aβ and AβT.

### Future Direction

Find out the pathways these miRNAs are involved in for further investigation of how they may contribute to the AD pathology.

### Limitation

Due to smaller input (less than 600 miRNAs) of the dataset, the DESeq2 normalization may not have been the most appropriate tool to assess this dataset. Therefore, I have also explored another package, NanoStringDiff (doi:%2010.1093/nar/gkw677). However, I have yet to figure out the appropriate design input to receive a result from this package (Error in Beta.full %\*% contrast : non-conformable arguments).

#### NanoStringDiff (Work in Progress)

\#\# Data Input

directory \<- "/Users/phoebechum/Desktop/NanoString/Data Analysis/R Analysis/Raw Data.csv"

designs = data.frame(group=c("MCI1","MCI2","MCI3","MCI4","MCI5","MCI6","YC1","YC2","YC3","YC4","YC5","YC6","AB1","AB2","AB3","AB4","AB5","AB6","ABT1","ABT2","ABT3","ABT4","ABT5","ABT6"))

library("NanoStringDiff")

NanoStringData=createNanoStringSetFromCsv(directory,header=TRUE,designs)

NanoStringData

pheno=pData(NanoStringData)

group=pheno\$group

design.full=model.matrix(\~0+group)

design.full

contrast = c(-1,1)

NanoStringData=estNormalizationFactors(NanoStringData)

positiveFactor(NanoStringData)

negativeFactor(NanoStringData)

housekeepingFactor(NanoStringData)

result=glm.LRT(NanoStringData,design.full,contrast=contrast)

head(result\$table)

result

\#\# Pairwise Comparisons

endogenous=matrix(rpois(300,50),25,12)

colnames(endogenous)=paste("Sample", 1:12)

colnames(endogenous)=paste("Sample",1:12)

positive=matrix(rpois(72,c(128,32,8,2,0.5,0.125)\*80),6,12)

negative=matrix(rpois(96,10),8,12)

housekeeping=matrix(rpois(36,100),3,12)

designs=data.frame(group=c(rep("YC",4),rep("MCI",4),rep("AB",4)),

+ gender=rep(c("Male","Male","Female","Female"),3),

+ age=c(20,40,39,37,29,47,23,45,34,65,35,64))

NanoStringData2=createNanoStringSet(endogenous,positive,

+ negative,housekeeping,designs)

NanoStringData2
